The Precambrian paleogeography of Laurentia

### 2020-09-24 Manuscript/Laurentia_Paleogeo_Manuscript.tex
implement edits across all sections

\subsection{Introduction and broad tectonic history}
\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly (Fig. \ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989c}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\textbf{Simplified map of the tectonic units of Laurentia.} The location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust are simplified from \cite{Whitmeyer2007a} with additions for Greenland based on \cite{St-Onge2009a}. Proterozoic orogens are labeled with \textit{italicized text} (or. -- orogen; THO  -- Trans-Hudson orogen; prov. -- province). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel \citep{Evans2021a}.}
\label{fig:Laurentia_map}
\end{figure*}

@@

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny (Fig. \ref{fig:tectonic_history}). Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to NE Canada in Labrador where is called the Labradorian orogeny that is interpreted to have been active from ca. 1.71 to 1.60 Ma \citep{Gower1992a, Gower2008b}. Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion. An interval of major plutonism occurred ca. 1.48 to 1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}. This plutonism is interpreted to be associated with a combination of continental arc magmatism and melt generation within a back-arc region Laurentia's long-live active margin \citep{Bickford2015a}. Metamorphic rocks from Mesoproterozoic sedimentary and volcanic protoliths in northern New Mexico have been interpreted to indicate an interval of ca. 1.46 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a, Aronoff2016a}. Younger magmatic activity ca. 1.37 Ga of the Southern Granite-Rhyolite Province suggests a similar active margin setting at that time \citep{Bickford2015a}. While an active margin interpretation for the Granite-Rhyolite Province with arc and back-arc magmatism in a back-arc setting has gained traction within the literature, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Tectonic_history.pdf}
\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown. The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown. Abbreviations on the figure: CAMP (Central Atlantic Magmatic Province); proto-Cord (proto-Cordilleran).}
\label{fig:tectonic_history}
\end{figure*}

Accretionary orogenesis continued along the eastern margin of Laurentia with the amalgamation and accretion of arcs associated with the ca. 1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the collision of terrane comprised of a previously rifted fragment of Laurentia and led to obduction of the Pyrites Complex ophiolite \citep{McLelland2010a, Chiarenzelli2011a}. The Shawinigan orogeny is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collisional orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. An exception to this quiescence during the interval between the Shawinigan and Grenvillian orogenies is ca. 1.15 to 1.12 Ga orogenesis in the Llano uplift of the southern Laurentia margin \citep{Mosher1998a}. Llano orogenesis is interpreted to have resulted from collision of continental lithosphere along with an accreted arc \citep{Mosher1998a}. This orogenesis is earlier and temporally distinct from the Grenvillian orogeny, is only known from a limited spatial area, and is located in a region that experienced further orogenesis during the Grenvillian orogeny \citep{Grimes2004a}. Taken together, this context is suggestive of a microcontinent collision leading to Llano uplift orogenesis prior to terminal Grenvillian continental collision. If this interpretation is correct, it is similar to Paleozoic orogenesis along the margin where microcontinent collision resulted in the Acadian orogeny prior to Alleghanian orogenesis during the Appalachian orogenic interval (\ref{fig:tectonic_history}).

@@

In the latest Mesoproterozoic (ca. 1.11 to 1.08 Ga) prior to the Grenvillian orogeny, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}. This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior. Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}.

@@

This late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.16 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian collisional orogenesis \citep{Karlstrom2001a}. This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g. Amazonia in \citealp{Elming2009a, Elming2021a}). The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and (an)other continent(s) ca. 1080 Ma -- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}. This extensive and major collisional orogenic history recorded in Laurentia, and also present on other Proterozoic continents, at this time remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition. Note that while the term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage. These constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}).

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, rifting occurred ca. 780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}. However, this extensional basin development predates the rifting that led to well-documented passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a, Levy1991a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia). If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (post 635 Ma as in \citealp{Levy1991a} and \citealp{Witkosky2018a}). The geological evidence therefore supports prolonged active tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), but suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up. The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca. 780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard. In this model, subsequent continent rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

In northwest Laurentia from the Ogilvie Mountains of Yukon to Victoria Island, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}. Tectonic extension is recorded in units of the lower Fifteenmile Group with maximum depositional ages of ca. 1050 Ma with ongoing basin development ca. 812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) potentially through thermal subsidence \citep{Macdonald2012a}. Earlier basin development in the region recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age -- older than the Fifteenmile Group and younger than the ca. 1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}).

Another margin that experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig. \ref{fig:tectonic_history}). Available geochronological constraints and thermal subsidence modeling indicate ca. 850 to 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}. These data suggest that conjugate continental lithosphere had rifted away from northeast Greenland by ca. 820 Ma.

@@

Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic Eras provides strong evidence for mobile plate tectonics driving Laurentia's evolution throughout the past 2 billion years. This tectonic history inferred from geological data can be enhanced through integration with the paleomagnetic record.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_poles_combined.pdf}
\caption{\textbf{Paleomagnetic poles from Laurentia.} Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including Greenland, Scotland and Svalbard). Bottom panel: Paleomagnetic data for Archean Provinces prior to the amalgamation of Laurentia.}
\label{fig:Laurentia_poles}
\end{figure*}

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results \citep{Evans2021a}. While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland \citep{Brown2018a}. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenvillian orogen. Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions (Table 2). However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2). An additional not-rated pole included in the present compilation is the new pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwest Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2019a} are also included.

@@

Many different paleogeographic connections between Laurentia and other Proterozoic cratons that have been proposed and utilized in paleogeographic models both prior to and following the amalgamation of Laurentia's constituent Archean provinces to form Laurentia. This section does not seek to be comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models -- particularly those following the assembly of Laurentia.

>>>>

\subsection{Introduction and broad tectonic history}
\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly (Fig. \ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989c}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\textbf{Simplified map of the tectonic units of Laurentia.} The location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust are simplified from \cite{Whitmeyer2007a} with additions for Greenland based on \cite{St-Onge2009a}. Proterozoic orogens are labeled with \textit{italicized text} (or. -- orogen; THO  -- Trans-Hudson orogen; MCR -- Midcontinent Rift). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel \citep{Evans2021a}.}
\label{fig:Laurentia_map}
\end{figure*}

@@

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny (Fig. \ref{fig:tectonic_history}). Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to NE Canada in Labrador where it is called the Labradorian orogeny that is interpreted to have been active from ca. 1.71 to 1.60 Ma (Fig. \ref{fig:Laurentia_map}; \citealp{Gower1992a, Gower2008b}).

Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion. An interval of major plutonism occurred ca. 1.48 to 1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}. This plutonism is interpreted to be associated with a combination of continental arc magmatism and melt generation within a back-arc region Laurentia's long-live active margin \citep{Bickford2015a}. Metamorphic rocks from Mesoproterozoic sedimentary and volcanic protoliths in northern New Mexico have been interpreted to indicate an interval of ca. 1.46 to 1.40 Ga orogenesis that has been named the Picuris orogeny \citep{Daniel2013a, Aronoff2016a}. Younger magmatic activity (ca. 1.37 Ga) of the Southern Granite-Rhyolite Province suggests a similar active margin setting at that time \citep{Bickford2015a}. While an active margin interpretation for the Granite-Rhyolite Province, with arc and back-arc magmatism, has gained traction within the literature, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Tectonic_history.pdf}
\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown. The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown. Abbreviations on the figure: CAMP -- Central Atlantic Magmatic Province; proto-Cord -- proto-Cordilleran.}
\label{fig:tectonic_history}
\end{figure*}

Accretionary orogenesis continued along the eastern margin of Laurentia with the amalgamation and accretion of arcs associated with the ca. 1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the collision of terrane comprised of a previously rifted fragment of Laurentia and led to obduction of the Pyrites Complex ophiolite \citep{McLelland2010a, Chiarenzelli2011a}. The Shawinigan orogeny is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collisional orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. An exception to this quiescence during the interval between the Shawinigan and Grenvillian orogenies is ca. 1.15 to 1.12 Ga orogenesis in the Llano uplift of the southern Laurentia margin \citep{Mosher1998a}. Llano orogenesis is interpreted to have resulted from collision of continental lithosphere along with an accreted arc \citep{Mosher1998a}. This orogenesis is earlier and temporally distinct from the Grenvillian orogeny, is only known from a limited spatial area, and is located in a region that experienced further orogenesis during the Grenvillian orogeny \citep{Grimes2004a}. Taken together, this context is suggestive of a microcontinent collision leading to Llano uplift orogenesis prior to terminal Grenvillian continental collision. If this interpretation is correct, it is similar to Paleozoic orogenesis along the margin where microcontinent collision resulted in the Acadian orogeny prior to Alleghanian orogenesis during the Appalachian orogenic interval (Fig. \ref{fig:tectonic_history}).

@@

In the latest Mesoproterozoic (ca. 1.11 to 1.08 Ga) prior to the Grenvillian orogeny, a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and adjacent Paleoproterozoic provinces to the south \citep{Cannon1992b}. This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior. Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Cannon1994a, Swanson-Hysell2019a}.

@@

This late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.16 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian collisional orogenesis \citep{Karlstrom2001a}. This constraint is incorporated into paleogeographic models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g. Amazonia in \citealp{Elming2009a, Elming2021a}). The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and (an)other continent(s) ca. 1080 Ma --- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}. This extensive and major collisional orogenic history recorded in Laurentia, and also present on other Proterozoic continents, at this time remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition. Note that while the term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage. These constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}).

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, rifting occurred ca. 780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}. However, this extensional basin development predates the rifting that that is interpreted to have been followed by passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a, Levy1991a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia). If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (post 635 Ma as in \citealp{Levy1991a} and \citealp{Witkosky2018a}). The geological evidence therefore supports prolonged active tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), but suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up. The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca. 780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard. In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

In northwest Laurentia from the Ogilvie Mountains of Yukon to Victoria Island, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}. Tectonic extension is recorded in units of the lower Fifteenmile Group with maximum depositional ages of ca. 1050 Ma with ongoing basin development ca. 812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) potentially through thermal subsidence \citep{Macdonald2012a}. Earlier basin development in the region recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age --- older than the Fifteenmile Group and younger than the ca. 1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}).

Another margin that experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig. \ref{fig:tectonic_history}). Available geochronological constraints and thermal subsidence modeling indicate ca. 850 to 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}. These data suggest that conjugate continental lithosphere rifted away from northeast Greenland by ca. 820 Ma.

@@

Even without considering other continents on Earth, the geological record of Paleoproterozoic collision of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic Eras provides strong evidence for mobile plate tectonics driving Laurentia's evolution throughout the past 2 billion years. This tectonic history inferred from geological data can be enhanced through integration with the paleomagnetic record.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_poles_combined.pdf}
\caption{\textbf{Paleomagnetic poles from Laurentia.} Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including those from the separated terranes of Greenland, Scotland and Svalbard rotated to Laurentia coordinates). Bottom panel: Paleomagnetic data for Archean provinces prior to the amalgamation of Laurentia.}
\label{fig:Laurentia_poles}
\end{figure*}

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results \citep{Evans2021a}. While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland \citep{Brown2018a}. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a} in conjunction with the papers in which they were reported. Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenvillian orogen. Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions (Table 2). However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citealp{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2). An additional not-rated pole included in the present compilation is the new pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwest Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2019a}, incorporating data from \cite{Weil2004a}, are also included.

@@

Many different paleogeographic connections between Laurentia and other Proterozoic cratons have been proposed and utilized in paleogeographic models both prior to and following the amalgamation of Laurentia's constituent Archean provinces. This section is not comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models --- particularly those following the assembly of Laurentia.
